Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of a simulation model

### 2025-12-03 paper/paper.qmd
SD removed

Here, we describe an simulation model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods. The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}. We compare an \acr{ABM} and a stochastic \acr{SD} model, and explore the advantages and disadvantages of each approach for this application.

@@

## Stochastic system dynamics model

To build a high-level stochastic system dynamics model, the model was first described in terms of the following differential equations (see @fig-sd-model), an adaptation of *SI* (susceptible--infection) models [@andersonDiscussionKermackMcKendrickEpidemic1991; @keelingModelingInfectiousDiseases2011]:

$$
\frac{dS}{dt} = (1-\lambda)A - D \cdot \frac{S}{S + I} - S (\beta + r)
$$

$$
\frac{dI}{dt} = \lambda A - D \cdot \frac{I}{S + I} + S (\beta + r)
$$

-   $S$ is the number of susceptible inpatients,
-   $I$ is the number of infected inpatients,
-   $dS/dt$ and $dI/dt$ are the rates of change of susceptible/non-infected and infected inpatients, respectively,
-   $A$ is the number of new admissions,
-   $D$ is the number of patient discharges,
-   $\lambda$ is the community infection rate,
-   $\beta$ is the transmission coefficient,
-   $r$ is the rate of random (*de novo*) infection.

The hospital capacity was constantly kept at 100% occupancy (i.e., patients were immediately replaced on discharge), therefore, $d = a$.

```{r fig-sd-model, include=TRUE}
#| echo: false
#| fig.cap: "System dynamics model. $S$ and $I$ are the number of susceptible and infected patients, respectively; $A$ and $D$ are the rates of admitted and discharged patients, respectively; $\\lambda$ is the community infection rate; $\\beta$ is the transmission coefficient; and $r$ is the rate of random (*de novo*) infection."
knitr::include_graphics("./images/sd_model.svg")
```

Next, the number of susceptible and infected patients at each discrete time $t + 1$ was calculated by first determining the number of patients discharged:

$$
D_{\text{total}, t} \sim \text{Poisson}(d),
$$

where $d$ is the discharge rate and $D_{\text{total}, t}$ is the total number of patients (both susceptible, $D_S$, and infected $D_I$), discharged at time $t$.

Therefore, $D_{\text{total}, t} = D_{S,t} + D_{I,t}$, where:

$$
D_{S,t} = \left \lfloor D_{\text{total}, t} \cdot \frac{S_t}{S_t + I_t} \right \rceil,
$$

and

$$
D_{I,t} = \left \lfloor D_{\text{total}, t} \cdot \frac{I_t}{S_t + I_t} \right \rceil = D_{\text{total}, t} - D_{S,t},,
$$

where $\lfloor x \rceil$ is the rounding of $x$ to the nearest integer.

Next, new admissions were calculated. Since $A_{\text{total}, t} = D_{\text{total}, t}$,

$$
A_{I,t} \sim \text{Binomial}(A_{\text{total}, t}, \lambda),
$$

and

$$
A_{S,t} = A_{\text{total}, t} - A_{I,t},
$$

To calculate the number of hospital-acquired infections (i.e., $\beta S_t + r S_t$), the binomial distribution was used:

$$
\text{HAI}_{\text{transmission}, t} \sim \text{Binomial}(S_t - D_{S, t}, \beta),
$$

and

$$
\text{HAI}_{\text{de novo}, t} \sim \text{Binomial}(S_t - D_{S, t} - \text{HAI}_{\text{transmission}, t}, r),
$$

where $\text{HAI}_{\text{transmission}, t}$ is the number of hospital-acquired infections due to transmission at time $t$, and $\text{HAI}_{\text{de novo}, t}$ is the number of hospital-acquired infections due to *de novo* infection at time $t$.

Finally, the number of susceptible and infected patients at time $t + 1$ was updated as:

$$
S_{t+1} = S_t + A_{S,t} - D_{S,t} - \text{HAI}_{\text{transmission}, t} - \text{HAI}_{\text{de novo}, t},
$$

and

$$
I_{t+1} = I_t + A_{I,t} - D_{I,t} + \text{HAI}_{\text{transmission}, t} + \text{HAI}_{\text{de novo}, t}.
$$

@@

To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures.

A similar approach was used to generate outbreaks in the stochastic system dynamics model. To align with the agent-based model, parameters were changed at four discrete time points: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures. Three parameter regimes were used: baseline (pre- and post-outbreak), outbreak (pre-control), and control (during enhanced IPC). During the control period only the transmission coefficient $\\beta$ changed; the community infection rate $\\lambda$ was unaffected by control measures (it only differed between baseline and outbreak). The parameter $r$ (rate of random infection) was kept constant throughout the simulation.

@@

### Calibration of \acr[first_use=TRUE]{ABM}

@@

Otherwise, $\theta_i$ was rejected. The parameter samples that generated the accepted simulations were used to approximate a posterior distribution. $\epsilon$ was set such that the acceptance rate was approximately 2.5%, which was deemed to be a good compromise between the accuracy of the posterior distribution and the computational cost of running the simulations.

@@

As a posterior predictve check, the posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios that were compared to the observed rates. To sample from the posterior distribution (i.e., the parameter sets retained after the rejection ABC algorithm) were samples with replacement were taken and used as an input to the stochastic \acr{SD} model.

@@

### Calibration of \acr[first_use=TRUE]{SD} model

A similar calibration approach using \acr{ABC} was also used to calibrate the stochastic \acr{SD} model to the observed outbreak data, although the calibration was performed in a single step that encompassed the entire period of the observed data (i.e., pre-outbreak, outbreak, and control periods).

@@

Both the \acr{ABM} and the \acr{SD} model were able to simulate the faecal-oral transmission of infection in a hospital setting. Although the \acr{SD} was computationally simpler, and produced simulated infection rates that were very similar to the observed rates (see @fig-posterior-predictive-check), the \acr{ABM} was based on a simulation that more closely resembled the workings of a real hospital. Importantly, the parameters of the \acr{ABM} were directly based on modifiable infection control practices, such as toilet cleaning frequency and antimicrobial prescription rates. Therefore, the main analyses and senstivity analyses were performed using the \acr{ABM}.

>>>>

Here, we describe an simulation model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods. The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}.

@@

@@

To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures. By changing parameters, the model simulated the conditions that led to increased transmission of infection, followed by \acr{IPC} measures that reduced transmission.

@@

@@

Otherwise, $\theta_i$ was rejected, giving a set of accepted parameters $\hat{\theta}_1, \ldots, \hat{\theta}_m$, where $m \leq n$. The parameter samples---i.e., $\hat{\theta}$ that generated the accepted simulations were used to approximate a posterior distribution. $\epsilon$ was set such that the acceptance rate was approximately 2.5%, which was deemed to be a good compromise between the accuracy of the posterior distribution and the computational cost of running the simulations.

@@

As a posterior predictive check, the posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios that were compared to the observed rates. A set of parameters was generated by sampling with replacement from $\hat{\theta}$, and used as an input to the simulation model. The results of these simulations were compared to the observed data to assess how well the calibrated model could reproduce the observed outbreak dynamics.

@@

@@
